Generative Adversarial Networks and Gravitational Waves

### 2020-03-30 Paper_draft/main.tex
modified:   .DS_Store# 	new file:   Paper_draft/figures/adding.png 	new file:   Paper_draft/iopart-num.bib 	new file:   Paper_draft/iopart-num.bst 	new file:   Paper_draft/iopart-num_custom.bst 	modified:   Paper_draft/main.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\usepackage{xcolor}
\usepackage{pifont}
\usepackage{comment}
%\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows, calc}
\expandafter\let\csname equation*\endcsname\relax 
\expandafter\let\csname endequation*\endcsname\relax 
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\newcommand{\jordan}[1]{\textbf{\textcolor{red}{JORDAN: #1}}}
\newcommand{\siong}[1]{\textbf{\textcolor{blue}{SIONG: #1}}}
\newcommand{\chris}[1]{\textbf{\textcolor{green}{CHRIS: #1}}}
\newcommand{\michael}[1]{\textbf{\textcolor{orange}{MICHAEL: #1}}}
\begin{document}

@@

Gravitational-wave (GW) astronomy is now an established field, starting with the first detection of a binary black hole merger \cite{} on September 2015. Following this, the first and second observations runs (O1 and O2) of Advanced LIGO and Advanced Virgo reported several more mergers \cite{}. On August 2017 a binary neutron star merger was observed alongside its electron-magnetic counterpart for the first time, giving rise to multimessenger gravitational wave astronomy. 
GW bursts are transient signals of typically short duration ($<$ 1s) whose waveforms are not accurately modelled or are complex to re-produce. Astrophysical sources for such transients include: Core collapse supernova, Neutron star instabilities, Fallback accretion onto a neutron star, Non axisymetric deformation in magnetars, Pulsar glitches, Neutron star post-mergers.
Black hole accretion disk fragmentation.. Since GW bursts are un-modelled they are not sensitive to template based detection schemes such as matched-filtering \cite{}, instead, detection involves distinguishing the signal from detector noise. This is only possible if the detector noise is well characterised and the candidate signal can be differentiated from system or environmental glitches. As such, GW burst searches rely on an astrophysical burst signature appearing in multiple detectors.
Many GW burst algorithms \cite{} are tested and tuned using model waveforms that may or may not have astrophysical significance but have easy to define parameters and share characteristics of real bursts that is enough to simulate coincident non-stationary deviations between detectors. Such waveforms may have long-duration, short bandwidth (ringdowns), long-duration, large bandwidth (inspirals) and many algorithms make use of sine-Gaussians: a Gaussian modulated sine wave that is characterised by it's central frequency and narrow bandwidth. This makes it a great tool for diagnosing LIGOs sensitivity to frequency. 
We aim to explore the use of machine learning in generating and interpreting these mock GW burst signals. Neural networks have shown to replicate the sensitivities of matched filtering in GW detection \cite{} and rapid parameter estimation \cite{}, however these methods have primarily been focused on binary black hole signals and have not yet expanded to burst examples.

@@

A subset of deep learning that has seen fruitful development in recent years \cite{} is Generative Adversarial Networks (GANs). These unsupervised algorithms learn patterns in a given training data set using an adversarial process. The generations from GANs are state-of-the-art in fields such as high quality image fidelity, image and text-to-image translation and video prediciton \cite{} and there is notable works on time series generations /cite. 
GANs work to train two competing neural networks, consisting of a discriminator  that is trained to distinguish
between real and fake data and a generator that produces synthetic
reproductions of the real data. The generator perform a mapping
from an input noise vector \textbf{z} to its representation of the data and discriminator  maps its
input \textbf{x} to a probability that the input came form either the training
data or generator.  During training, discriminator  is given a batch of samples that contains one half real data (labelled as 1)
and one half fake data (labelled as 0) which it then makes predictions on. The
loss for discriminator  is calculated by comparing its predictions to the labelled data through the binary cross-entropy function. The training process
of a GAN alternatively updates the weights of the discriminator  and generator based on information
on its competitors loss function. This loss of discriminator  is used to update the weights
of generator to produce more realistic samples of the input distribution, the loss of G
encourages discriminator  to update its classification abilities. Both networks compete
in a minimax game that is tied to a value function V (D, G) which generator is trying
minimise and discriminator  is trying to maximise:

\begin{equation}
\mathop{\text{min}}_{G}  \mathop{\text{max}}_{D} V(D,G) = \mathbb{E}_{\mathbf{x} \sim p_{\text{r}}(\mathbf{x})} [\text{log} D(\mathbf{x})] \\ + \mathbb{E}_{\mathbf{z} \sim p_{\text{z}}(\mathbf{z})} [\text{log}(1-D(G(\mathbf{z})))]
\label{equation:GANloss}
\end{equation}
\subsection{Auxiliary conditional GANs}
In theory, the adversarial process will eventually lead to the local Nash equilibrium [18] where
both neural networks are trained optimally. In practice, however, GANs are
notoriously difficult to train. Such difficulties include: Non-convergence,
where the model parameters oscillate and the loss never converges, mode
collapse where G produces a limited diversity of samples, and diminishing
gradients when applying gradient descent to a
non-continuous function. To overcome some of these difficulties [blank]
proposed adding structure to the latent space by providing G with a class
label. This has the effect of making a point in latent space conditional on a
provided class. [blank] extends this idea further by requiring D to output a
probability of the data belonging to each class. A pictorial representation on the differences between these approaches is show below.

\begin{figure}[h!]
    \centering
    %\begin{subfigure}[h]{0.5\textwidth}
        %\centering
        \input{figures/GAN_diagram_vertical}%     without .tex extension
    %\end{subfigure}\hfill
    %\begin{subfigure}[h]{0.5\textwidth}
       % \centering
        \input{figures/ACGAN_diagram_vertical}%     without .tex extension
    %\end{subfigure}\hfill
    \caption{Comparison of the original GAN method and the Auxiliary Conditional-GAN method. For ACGANs the training data requires a label denoting its class that is also fed to the generator which then learns to generate waveforms based on the input label. Additionaly, the discrminator learns to classify which class the signal belongs to.}
\end{figure}

@@

\subsection{Vector Arithmetic}
\subsection{Classifier}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\bibliographystyle{iopart-num_custom}
\usepackage{xcolor}
\usepackage{pifont}
\usepackage{comment}
%\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows, calc}
\expandafter\let\csname equation*\endcsname\relax 
\expandafter\let\csname endequation*\endcsname\relax 
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\newcommand{\jordan}[1]{\textbf{\textcolor{red}{JORDAN: #1}}}
\newcommand{\siong}[1]{\textbf{\textcolor{blue}{SIONG: #1}}}
\newcommand{\chris}[1]{\textbf{\textcolor{green}{CHRIS: #1}}}
\newcommand{\michael}[1]{\textbf{\textcolor{orange}{MICHAEL: #1}}}
\begin{document}

@@

Gravitational-wave (GW) astronomy is now an established field, starting with the first detection of a binary black hole merger \cite{Abbott2016} on September 2015. Following this, the first and second observations runs (O1 and O2) of Advanced LIGO and Advanced Virgo reported several more mergers \cite{Abbott2016a, Abbott2017, Abbott2017a, Abbott2017b}. On August 2017 a binary neutron star merger was observed alongside its electron-magnetic counterpart for the first time, giving rise to multimessenger gravitational wave astronomy. 
GW bursts are transient signals of typically short duration ($<$ 1s) whose waveforms are not accurately modelled or are complex to re-produce. Astrophysical sources for such transients include: Core collapse supernova, Neutron star instabilities, Fallback accretion onto a neutron star, Non axisymetric deformation in magnetars, Pulsar glitches, Neutron star post-mergers.
As GW bursts are un-modelled they are not sensitive to template based detection schemes such as matched-filtering \cite{Owen1998}, instead, detection involves distinguishing the signal from detector noise. This is only possible if the detector noise is well characterised and the candidate signal can be differentiated from system or environmental glitches. As such, GW burst searches rely on an astrophysical burst signature appearing in multiple detectors.
Many GW burst algorithms \cite{Klimenko_2008, Aso_2008} are tested and tuned using model waveforms that may or may not have astrophysical significance but have easy to define parameters and share characteristics of real bursts that is enough to simulate coincident non-stationary deviations between detectors. Such waveforms may have long-duration, short bandwidth (ringdowns), long-duration, large bandwidth (inspirals) and many algorithms make use of sine-Gaussians: a Gaussian modulated sine wave that is characterised by it's central frequency and narrow bandwidth. This makes it a great tool for diagnosing LIGOs sensitivity to frequency. 
We aim to explore the use of machine learning in generating and interpreting these mock GW burst signals. Neural networks have shown to replicate the sensitivities of matched filtering in GW detection \cite{Gabbard2017} and rapid parameter estimation \cite{gabbard2019bayesian}, however these methods have primarily been focused on binary black hole signals and have not yet expanded to burst examples.

@@

A subset of deep learning that has seen fruitful development in recent years \cite{Goodfellow2014} is Generative Adversarial Networks (GANs). These unsupervised algorithms learn patterns in a given training data set using an adversarial process. The generations from GANs are state-of-the-art in fields such as high quality image fidelity \cite{brock2018large,karras2019analyzing}, text-to-image translation \cite{reed2016generative} and video prediction \cite{liang2017dual} as well as time series generations \cite{esteban2017realvalued}. 
GANs train two competing neural networks, consisting of a discriminator that is set up to distinguish
between real and fake data and a generator that produces synthetic
reproductions of the real data. The generator performs a mapping
from an input noise vector \textbf{z} to its representation of the data and the discriminator  maps its
input \textbf{x} to a probability that the input came form either the training
data or generator.  During training, the discriminator is given a batch of samples that contains one half real data
and one half fake data which it then makes predictions on. The
loss for discriminator  is calculated by comparing its predictions to the labelled data through the binary cross-entropy function. The training process
of a GAN alternatively updates the weights of the discriminator  and generator based on information
on its competitors loss function. This loss of discriminator  is used to update the weights
of generator to produce more realistic samples of the input distribution, the loss of G
encourages discriminator  to update its classification abilities. Both networks compete
in a minimax game which generator is trying
minimise and the discriminator is trying to maximise:

\begin{equation}
\mathop{\text{min}}_{G}  \mathop{\text{max}}_{D} V(D,G) = \mathbb{E}_{\mathbf{x} \sim p_{\text{r}}(\mathbf{x})} [\text{log} D(\mathbf{x})] \\ + \mathbb{E}_{\mathbf{z} \sim p_{\text{z}}(\mathbf{z})} [\text{log}(1-D(G(\mathbf{z})))]
\label{equation:GANloss}
\end{equation}
\subsection{Auxiliary conditional GANs}
In theory, the adversarial process will eventually lead to the local Nash equilibrium \cite{Nash1950} whereby
both neural networks are trained optimally. In practice, however, GANs are
notoriously difficult to train. Such difficulties include: Non-convergence,
where the model parameters oscillate and the loss never converges, mode
collapse where G produces a limited diversity of samples, and diminishing
gradients when applying gradient descent to a
non-continuous function. To overcome some of these difficulties, a conditional-GAN (CGANs) \cite{cgan}
adds structure to the latent space by providing the generator with a class or attribute
label. This has the effect of making a point in latent space conditional on a
provided class. This idea was extended further with auxiliary conditional GANs (ACGANs) \cite{odena2016conditional} that require the discriminator to output a
probability of data belonging to each class. A pictorial representation on the differences between these approaches is show below.

\begin{figure}
    \centering
        \input{figures/GAN_diagram_vertical}%     without .tex extension
        \input{figures/ACGAN_diagram_vertical}%     without .tex extension
    \caption{Comparison of the original GAN method and the Auxiliary Conditional-GAN method. For ACGANs the training data requires a label denoting its class that is also fed to the generator which then learns to generate waveforms based on the input label. Additionaly, the discrminator learns to classify which class the signal belongs to.}
\end{figure}

@@

\subsection{Vector Arithmetic}
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/adding.png}
    \caption{Vector arithmetic for unmodelled waveforms. Keeping the z vectors constant and generating two classes of signals, their class embedding vectors can the be summed to give a new waveform that is comprised of the two classes.}
    \label{fig:add}
\end{figure}
\subsection{Classifier}

@@

\section*{References}
\bibliography{iopart-num}
